Estimating duration in the presence of misclassification

### 2024-08-01 ATACCC.tex
Christl p73 minor comment
improve figure appearance
Christl p75 comment

$\tau_{\max,i}$ is a fixed effect for each individual because time is defined for each individual based on when they are first tested.
This time is based on logistical and behavioural factors such as delays in the NHS Test and Trace system and the study design.
\begin{figure}[tbhp]
  \centering \includegraphics{ATACCC/typical_trajectory}
  \caption[Example viral load model trajectory.]{%
    Example viral load trajectory produced by \textcite{hakkiOnset}'s viral load model, using realistic parameter values.
    The increase and decrease in log viral load are approximately linear with a gradient of $a_i$ and $b_i$ respectively.
    The peak is at time $\tau_{\max,i} = 0$ and log viral load $\log v_{\max,i}$.
    The key omits subscripts for brevity.
  }
  \label{ATACCC:fig:viral-load-model}
\end{figure}

@@

\textcite{hakkiOnset} uses weakly informative priors, similar to \textcite{singanayagamCommunity}, see \cref{ATACCC:table:hakki-priors}.
I reran \textcite{hakkiOnset}'s analysis using R~4.3.1~\autocite{R-4-3-1}.
The inference was performed using RStan~2.32.3~\autocite{RStan-2-32-3} and Stan~2.32.2~\autocite{Stan-2-32-2}.
Inference settings were the same as \textcite{hakkiOnset}, which included 8 chains with a total of \numprint{8000} iterations each with the first \numprint{3000} used as warm-up.
\begin{table}
\begin{tabular}{l c l l}
    Parameter description & Symbol & Distribution \\
    \hline \\
    Mean max log viral load & $\mu_{v_\text{max}, g}$ & $\N(15, 15^2)$ \\
    Standard deviation of max log viral load & $\sigma_{v_\text{max}}$ & $\N_+(0, 10^2)$ \\
    Mean rate of log viral load increase & $\mu_{a, g}$ & $\N(1.25, 0.75^2)$ \\
    Standard deviation of rate of log viral load increase & $\sigma_{a}$ & $\N_+(0, 1)$ \\
    Mean rate of log viral load decrease & $\mu_{b, g}$ & $\N(0.5, 1.4^2)$ \\
    Standard deviation of rate of log viral load decrease & $\sigma_{b}$ & $\N_+(0, 1)$ \\
    Correlation of viral load parameters & $\matr{C}_\phi$ & $\LKJ(1)$ \\
    Negative log of rate of test failure & $-\log \rho$ & $\N(3, 0.5)$ \\
    Mean log viral load for test failures & $x_0$ & $\N(0, 3^2)$ \\
    Standard deviation of log viral load for test failures & $\sigma_0$ & $\N_+(3, 3^2)$ \\
    Standard deviation of log viral load observations & $\sigma_v$ & $1 + \N_+(2, 3^2)$ \\
    Time from first test to peak & $\tau_{\max,i}$ & $\N(5,5)$
\end{tabular}
\caption[Viral load model priors]{Priors for each parameter, taken from \textcite{hakkiOnset}. $\N_+$ indicates a normal distribution truncated to the non-negative reals. Details of the distributions, including their parameterizations, are in \cref{E-distributions}. The elements of $\vec{\mu}_{\phi,g}$ and $\vec{\sigma}_\phi$ have independent priors placed on them, as specified in the first lines of the table. Priors are the same for all $g$ and $i$. \label{ATACCC:table:hakki-priors}}
\end{table}

@@

The original \textcite{hakkiOnset} model has convergence issues (Rhat > 1.07 and ESS (Effective Sample Size) < 100) for the parameters of the test failure distribution, $\sigma_0$ and $\rho$.
The simplified model I proposed in \cref{ATACCC:sec:observation-modification} solved this issue, meaning that this analysis converged.

The two models give very similar estimates for the population-level parameters (see \cref{ATACCC:fig:compare-hakki-modified}).
These are the main parameters of interest (as explained in \cref{ATACCC:sec:duration}), suggesting that the results are robust to the choice of model.
As the estimates are similar and the modified model converges better, I will use only the results from the modified model.
In this study, $\vlod = 0$.
\begin{figure}
  \centering \includegraphics{ATACCC/compare_hakki_modified}
  \caption[Comparison of population-level parameters between models.]{Kernel-density smoothed posterior densities for the population mean (top), $\vec{\mu}_{\phi,1}$, and covariance matrix (bottom), $\matr{\Sigma}_\phi$, for both the \textcite{hakkiOnset} model (described in \cref{ATACCC:sec:hakki}) and the model with a modified observation model to improve convergence (described in \cref{ATACCC:sec:observation-modification}). Mean only shown for the unvaccinated group, the primary group of interest. The groups share the covariance matrix. These parameters, which are the primary ones of interest (see \cref{ATACCC:sec:duration}), have very similar estimates between the models. \label{ATACCC:fig:compare-hakki-modified}}
\end{figure}

Clear differences emerge between the vaccinated and unvaccinated groups (see \cref{ATACCC:fig:median-trajectories} and \cref{ATACCC:table:population-estimates}).
The rates of viral load increase at similar rates in the two groups, the vaccinated group have higher peak viral loads and clear the virus faster.
This leads to a shorter duration of RT-PCR positivity in the vaccinated group.
\begin{figure}
  \centering \includegraphics{ATACCC/mean_trajectories}
  \caption[Median viral load trajectories]{Trajectory (posterior median and 95\% CrI) using the population median parameter values, $\exp \vec{\mu}_{\phi,g}$, excluding the within-population variation. These assume $\tau_{\max,i}=0$, however modifying this simply translates the trajectories left or right. \label{ATACCC:fig:median-trajectories}}
\end{figure}
\begin{table}
  \makebox[\linewidth]{
  \begin{tabular}{lcc}
    Parameter & Unvaccinated population & Vaccinated population \\
    Max log viral load, $\mu_{v_\text{max},g}$ & 14.9 (14.2--15.7) & 16.0 (15.2--16.8) \\
    Rate of log viral load increase, $\exp(\mu_{a,g})$ & 5.07 (3.70--7.06) & 1.25 (1.18--1.32) \\
    Rate of log viral load decrease, $\exp(\mu_{b,g})$ & 1.33 (1.09--1.63) & 1.89 (1.50--2.36) \\
  \end{tabular}
  }
  \caption{Population-level estimates of the viral load parameters (posterior medians and 95\% CrI). The vaccinated population have higher peak viral loads but clear the virus faster. \label{ATACCC:table:population-estimates}}
\end{table}

@@

The $F_{D,g}$ estimates for the two groups are in \cref{ATACCC:fig:duration}.
The vaccinated population very likely has a shorter duration of RT-PCR positive (93\% posterior probability of a shorter median duration) than the unvaccinated population (see \cref{ATACCC:fig:duration}).
The median duration of RT-PCR positivity for the vaccinated population is 13 days (95\% CrI: 11--16 days), but it is 16 days (95\% CrI: 14--19 days) for the unvaccinated population.
\begin{figure}
  \centering \includegraphics{ATACCC/duration}
  \caption[Duration of RT-PCR positivity.]{Estimated distribution (survival function) of duration of RT-PCR positivity for the two populations. The vaccinated population have shorter durations of RT-PCR positivity.  \label{ATACCC:fig:duration}}
\end{figure}

The modified model produced good fits to the data (see \cref{ATACCC:fig:goodness-of-fits,ATACCC:fig:appendix-goodness-of-fits}).
Almost all test results are captured in the 95\% credible intervals.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{ATACCC/fits}}
  \caption[Posterior predictive viral load fits]{Posterior predictive observed viral loads (median and 95\% CrI, lines and ribbon) for selected individuals, overlaid with their observed test results (dots). Posterior estimates are conditional on the result not being a false negative. Negative RT-PCR tests are plotted at the limit of detection, \ie a log viral load of 0. Each plot is labelled with the individual's position in the dataset, an arbitrary value.
  Remaining individuals shown in \cref{ATACCC:fig:appendix-goodness-of-fits}.
  \label{ATACCC:fig:goodness-of-fits}}
\end{figure}

@@

Individual 55 has an extremely noisy set of observations (see \cref{ATACCC:fig:goodness-of-fits}).
This means that two viral load trajectories fit the data well (shown in \cref{ATACCC:fig:individual-55}).
The first, found by two chains, increases very quickly, peaks 4.0 days after the first test, then declines slowly.
The second, found by the other six chains, is more symmetrical, peaking 7.2 days after the first test.
The population-level parameter estimates agree between the two sets of chains (not shown); therefore, the issue should not affect any estimates of interest.
\begin{figure}
  \centering \includegraphics{ATACCC/fit_individual_55}
  \caption[Goodness-of-fit for individual 55]{Fit to individual 55's data coloured by MCMC chain. Two chains favour the quickly increasing viral load trajectory, peaking at day 4.0, the other six favour the slower trajectory, peaking at day 7.2. \label{ATACCC:fig:individual-55}}
\end{figure}

>>>>

$\tau_{\max,i}$ is a fixed effect for each individual because time is defined for each individual based on when they are first tested.
This time is based on logistical and behavioural factors such as delays in the NHS Test and Trace system and the study design.
\begin{figure}[tbhp]
  \makebox[\textwidth][c]{\includegraphics{ATACCC/typical_trajectory}}
  \caption[Example viral load model trajectory.]{%
    Example viral load trajectory produced by \textcite{hakkiOnset}'s viral load model, using realistic parameter values.
    The increase and decrease in log viral load are approximately linear with a gradient of $a_i$ and $b_i$ respectively.
    The peak is at time $\tau_{\max,i} = 0$ and log viral load $\log v_{\max,i}$.
    The key omits subscripts for brevity.
  }
  \label{ATACCC:fig:viral-load-model}
\end{figure}

@@

\textcite{hakkiOnset} uses weakly informative priors, similar to \textcite{singanayagamCommunity}, see \cref{ATACCC:table:hakki-priors}.
I reran \textcite{hakkiOnset}'s analysis using R~4.3.1~\autocite{R-4-3-1}.
The inference was performed using RStan~2.32.3~\autocite{RStan-2-32-3} and Stan~2.32.2~\autocite{Stan-2-32-2}.
Inference settings were the same as \textcite{hakkiOnset}, which included 8 chains with a total of \numprint{8000} iterations each with the first \numprint{3000} used as warm-up.
\begin{table}
\begin{tabular}{l c l l}
    Parameter description & Symbol & Distribution \\
    \hline \\
    Mean max log viral load & $\mu_{v_\text{max}, g}$ & $\N(15, 15^2)$ \\
    Standard deviation of max log viral load & $\sigma_{v_\text{max}}$ & $\N_+(0, 10^2)$ \\
    Mean rate of log viral load increase & $\mu_{a, g}$ & $\N(1.25, 0.75^2)$ \\
    Standard deviation of rate of log viral load increase & $\sigma_{a}$ & $\N_+(0, 1)$ \\
    Mean rate of log viral load decrease & $\mu_{b, g}$ & $\N(0.5, 1.4^2)$ \\
    Standard deviation of rate of log viral load decrease & $\sigma_{b}$ & $\N_+(0, 1)$ \\
    Correlation of viral load parameters & $\matr{C}_\phi$ & $\LKJ(1)$ \\
    Negative log of rate of test failure & $-\log \rho$ & $\N(3, 0.5)$ \\
    Mean log viral load for test failures & $x_0$ & $\N(0, 3^2)$ \\
    Standard deviation of log viral load for test failures & $\sigma_0$ & $\N_+(3, 3^2)$ \\
    Standard deviation of log viral load observations & $\sigma_v$ & $1 + \N_+(2, 3^2)$ \\
    Time from first test to peak & $\tau_{\max,i}$ & $\N(5,5)$
\end{tabular}
\caption[Viral load model priors]{Priors for each parameter, taken from \textcite{hakkiOnset}. $\N_+$ indicates a normal distribution truncated to the non-negative reals. Details of all distributions used, including their parameterizations, are in \cref{E-distributions}. The elements of $\vec{\mu}_{\phi,g}$ and $\vec{\sigma}_\phi$ have independent priors placed on them, as specified in the first lines of the table. Priors are the same for all $g$ and $i$. \label{ATACCC:table:hakki-priors}}
\end{table}

@@

The original \textcite{hakkiOnset} model has convergence issues (Rhat > 1.07 and ESS (Effective Sample Size) < 100) for the parameters of the test failure distribution, $\sigma_0$ and $\rho$; typically Rhat < 1.05 and ESS of at least a few hundred are recommended to ensure that the inference procedure has converged to the true posterior distribution with adequate precision (see \cref{E-MCMC}).
The simplified model I proposed in \cref{ATACCC:sec:observation-modification} solved this issue, meaning that this analysis converged.

The two models give very similar estimates for the population-level parameters (see \cref{ATACCC:fig:compare-hakki-modified}).
These are the main parameters of interest (as explained in \cref{ATACCC:sec:duration}), suggesting that the results are robust to the choice of model.
As the estimates are similar and the modified model converges better, I will use only the results from the modified model.
In this study, $\vlod = 0$.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics{ATACCC/compare_hakki_modified}}
  \caption[Comparison of population-level parameters between models.]{Kernel-density smoothed posterior densities for the population mean (top), $\vec{\mu}_{\phi,1}$, and covariance matrix (bottom), $\matr{\Sigma}_\phi$, for both the \textcite{hakkiOnset} model (described in \cref{ATACCC:sec:hakki}) and the model with a modified observation model to improve convergence (described in \cref{ATACCC:sec:observation-modification}). Mean only shown for the unvaccinated group, the primary group of interest. The groups share the covariance matrix. These parameters, which are the primary ones of interest (see \cref{ATACCC:sec:duration}), have very similar estimates between the models. \label{ATACCC:fig:compare-hakki-modified}}
\end{figure}

Clear differences emerge between the vaccinated and unvaccinated groups (see \cref{ATACCC:fig:median-trajectories} and \cref{ATACCC:table:population-estimates}).
The rates of viral load increase at similar rates in the two groups, the vaccinated group have higher peak viral loads and clear the virus faster.
This leads to a shorter duration of RT-PCR positivity in the vaccinated group.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics{ATACCC/mean_trajectories}}
  \caption[Median viral load trajectories]{Trajectory (posterior median and 95\% CrI) using the population median parameter values, $\exp \vec{\mu}_{\phi,g}$, excluding the within-population variation. These assume $\tau_{\max,i}=0$, however modifying this simply translates the trajectories left or right. \label{ATACCC:fig:median-trajectories}}
\end{figure}
\begin{table}
  \makebox[\linewidth]{
  \begin{tabular}{lcc}
    Parameter & Unvaccinated population & Vaccinated population \\
    Max log viral load, $\mu_{v_\text{max},g}$ & 14.9 (14.2--15.7) & 16.0 (15.2--16.8) \\
    Rate of log viral load increase, $\exp(\mu_{a,g})$ & 5.07 (3.70--7.06) & 1.25 (1.18--1.32) \\
    Rate of log viral load decrease, $\exp(\mu_{b,g})$ & 1.33 (1.09--1.63) & 1.89 (1.50--2.36) \\
  \end{tabular}
  }
  \caption{Population-level estimates of the viral load parameters (posterior medians and 95\% CrI). The vaccinated population have higher peak viral loads but clear the virus faster. \label{ATACCC:table:population-estimates}}
\end{table}

@@

The $F_{D,g}$ estimates for the two groups are in \cref{ATACCC:fig:duration}.
The vaccinated population very likely has a shorter duration of RT-PCR positive (93\% posterior probability of a shorter median duration) than the unvaccinated population (see \cref{ATACCC:fig:duration}).
The median duration of RT-PCR positivity for the vaccinated population is 13 days (95\% CrI: 11--16 days), but it is 16 days (95\% CrI: 14--19 days) for the unvaccinated population.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics{ATACCC/duration}}
  \caption[Duration of RT-PCR positivity.]{Estimated distribution (survival function) of duration of RT-PCR positivity for the two populations. The vaccinated population have shorter durations of RT-PCR positivity.  \label{ATACCC:fig:duration}}
\end{figure}

The modified model produced good fits to the data (see \cref{ATACCC:fig:goodness-of-fits,ATACCC:fig:appendix-goodness-of-fits}).
Almost all test results are captured in the 95\% credible intervals.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics{ATACCC/fits}}
  \caption[Posterior predictive viral load fits]{Posterior predictive observed viral loads (median and 95\% CrI, lines and ribbon) for selected individuals, overlaid with their observed test results (dots). Posterior estimates are conditional on the result not being a false negative. Negative RT-PCR tests are plotted at the limit of detection, \ie a log viral load of 0. Each plot is labelled with the individual's position in the dataset, an arbitrary value.
  Remaining individuals shown in \cref{ATACCC:fig:appendix-goodness-of-fits}.
  \label{ATACCC:fig:goodness-of-fits}}
\end{figure}

@@

Individual 55 has an extremely noisy set of observations (see \cref{ATACCC:fig:goodness-of-fits}).
This means that two viral load trajectories fit the data well (shown in \cref{ATACCC:fig:individual-55}).
The first, found by two chains, increases very quickly, peaks 4.0 days after the first test, then declines slowly.
The second, found by the other six chains, is more symmetrical, peaking 7.2 days after the first test.
The population-level parameter estimates agree between the two sets of chains (not shown); therefore, the issue should not affect any estimates of interest.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics{ATACCC/fit_individual_55}}
  \caption[Goodness-of-fit for individual 55]{Fit to individual 55's data coloured by MCMC chain. Two chains favour the quickly increasing viral load trajectory, peaking at day 4.0, the other six favour the slower trajectory, peaking at day 7.2. \label{ATACCC:fig:individual-55}}
\end{figure}
